---
title: "Dose-Response Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose-Response Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
library(cellreportR)
```

cellreportR fits three standard dose-response models:

- **4PL** (four-parameter log-logistic)
- **3PL** (fixed lower asymptote at zero)
- **Linear** (for convenience; fits `y ~ x`)

## Setting up a dose-response experiment

Any `cr_experiment` with a `dose` column in `design` is eligible.

```{r setup-doses}
exp <- cr_example_experiment(seed = 3, n_cells_per_well = 60)
exp$design$dose <- dplyr::case_when(
  exp$design$treatment == "Untreated"      ~ 0.1,
  exp$design$treatment == "CompoundA_low"  ~ 50,
  exp$design$treatment == "CompoundA_high" ~ 500,
  exp$design$treatment == "PosControl"     ~ 1000,
  TRUE ~ 10
)
```

## Fitting

```{r fit}
fit <- cr_dose_response(exp,
                        channel = "marker_1",
                        model = "4pl",
                        log_dose = TRUE)
fit$model
fit$params
```

## IC50 / EC50

```{r ic50}
cr_ic50(fit)
```

## Plotting

```{r plot-dr}
cr_plot_dose_response(fit)
```

## Troubleshooting

If `nls()` fails to converge (for example when there are very few
dose levels), cellreportR falls back to a linear model. Check
`fit$model` after fitting.
