---
title: "Publication-ready figures with cellreportR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publication-ready figures with cellreportR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=6,fig.height=3.6)
```

cellreportR uses position first, then shape or line type, then colour. Colour is
intentionally redundant for categorical distinctions. The restrained default
palette is designed with common colour-vision deficiencies in mind, without a
claim of formal accessibility certification.

```{r scales}
cellreportR::cr_palette()
cellreportR::cr_shapes(4)
cellreportR::cr_linetypes(4)
cellreportR::cr_plot_style("grayscale")
```

```{r estimates}
d <- data.frame(group=factor(c("Reference","Group A","Group B"),levels=c("Reference","Group A","Group B")),estimate=c(0.1,0.8,1.3),conf_low=c(-0.2,0.3,0.7),conf_high=c(0.4,1.3,1.9))
cellreportR::cr_plot_estimates(d,reference=0,colour_mode="colour")
cellreportR::cr_plot_estimates(d,reference=0,colour_mode="grayscale")
```

```{r result-position}
cellreportR::cr_plot_result_position(1.42,c(1,2),c("Lower","Intermediate","Upper"),unit="a.u.",classification="HIGH",mode="grayscale")
```

```{r qc}
qc <- cellreportR::cr_report_qc(c("Criterion A","Criterion B"),c("met","review"),c("Configured criteria","Configured criteria"),c("PASS","WARN"))
cellreportR::cr_plot_qc_summary(qc)
```

Exports have explicit dimensions and a white background, independent of an
interactive plotting pane.

```{r export, eval=FALSE}
cellreportR::cr_save_figure(p,"figure.pdf",size="single")
cellreportR::cr_save_figure(p,"figure-wide.pdf",size="double")
cellreportR::cr_save_figure(p,"figure.tiff",size="double",dpi=600)
```

The style layer does not filter, transform, reorder analytically meaningful
values, or select results. Factor levels are respected; explicit ordering is
available in estimate plots. More than the reliable number of colours, shapes,
or line types produces a warning so callers can prefer facets or direct labels.
