---
title: "Getting Started with cellreportR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cellreportR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4, fig.align = "center"
)
options(width = 84)
library(cellreportR)
```

cellreportR takes segmented single-cell microscopy data to a structured,
auditable set of estimates. This vignette walks the short path on one plate
using bundled synthetic data; `vignette("end-to-end")` walks the long one,
from files on disk to a finished report.

## 1. Load example data

`cr_example_experiment()` simulates a 96-unit plate: six treatment levels,
four channels, two plates and two pre-treatment intervals. Sub-threshold
debris, a plate-edge effect and a pair of saturated units are built in, so the
quality-control functions have something real to remove.

```{r example}
exp <- cr_example_experiment(seed = 42, n_cells_per_well = 60)
exp
```

## 2. Inspect the design

```{r design}
head(exp$design)
```

The unit of replication is the well, not the cell. `cr_n_cells()` counts by
any design column:

```{r counts}
cr_n_cells(exp, by = "treatment")
```

A batch here is the combination of two columns, which is what later
standardization references:

```{r batch}
exp$batch_vars
table(exp$design$plate, exp$design$interval)
```

## 3. Apply quality control

```{r qc}
exp_qc <- exp |>
  cr_qc_filter(min_area = 50, max_area = 5000, min_circularity = 0.2) |>
  cr_qc_doublets(k = 2.5)

cr_qc_summary(exp_qc)[, c("step", "cells_before", "cells_after",
                          "percent_removed")]
```

The QC log is append-only and travels with the object, so what was removed
stays recoverable after the fact.

## 4. Fold changes and tests

`cr_test_all()` contrasts every treatment level against the control and
attaches a summary table with the multiplicity adjustments alongside — never
instead of — the unadjusted p-value.

```{r analysis}
res <- cr_test_all(exp_qc,
                   channel       = "marker_1",
                   control_group = "Untreated",
                   level         = "replicate")

attr(res, "summary")[, c("treatment", "log2_fc", "p_value", "p_BH",
                         "cohens_d", "interpretation")]
```

`level = "replicate"` aggregates cells to one value per well before testing.
That is the defensible default; `vignette("statistical-analysis")` shows what
the cell-level alternative costs.

## 5. Visualize

```{r plots}
cr_plot_intensity(exp_qc, "marker_1")
cr_plot_effect_sizes(res)
```

```{r plate-map}
cr_plot_plate(exp_qc, "marker_1", metric = "median")
```

## 6. Discriminability

When the question is whether a marker separates treated from control cells
rather than whether the group means differ, fit a univariate logistic
regression and read the area under the ROC curve:

```{r logistic}
logit <- cr_logistic(exp_qc,
                     channel   = "marker_1",
                     treatment = "CompoundA_high",
                     control   = "Untreated")
cr_auc(logit)
cr_plot_roc(logit)
```

## 7. Next steps

* `vignette("end-to-end")` — ingest, units, the quality-control gate,
  per-batch standardization, effect sizes at two levels, sample size and the
  report.
* `vignette("statistical-analysis")` — cell-level versus replicate-level
  inference.
* `vignette("dose-response")` — fitting exposure-response curves.
* `cr_run_app()` — the same pipeline in a browser; see
  `vignette("shiny-app")`.

# Use of LLM tools

Portions of this package were prepared with assistance from large language model tooling for
narrowly defined, non-authorial tasks: copyediting, prose smoothing, Markdown/LaTeX formatting,
scaffolding of boilerplate files (CI configs, build scripts), code refactoring. The tools used were [Chat AI](https://kisski.gwdg.de/leistungen/2-02-llm-service/),
the LLM service of KISSKI (GWDG), and a self-hosted **Mistral Small (24B, Apache-2.0)** run locally via
[Ollama](https://ollama.com/) and the `ollamar` R package — local inference only, with no data sent to
third parties for the self-hosted model.

All scientific claims, methodological choices, analyses, interpretations, and conclusions are the
author's own. No LLM-generated text was incorporated without review and revision, and every reference
was verified against its DOI, arXiv ID, or ISBN.
