---
title: "Getting Started with cellreportR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cellreportR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-heller/cellreportR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-heller/cellreportR/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/r-heller/cellreportR/actions/workflows/pkgdown.yaml/badge.svg)](https://r-heller.github.io/cellreportR/)
[![CRAN status](https://www.r-pkg.org/badges/version/cellreportR)](https://CRAN.R-project.org/package=cellreportR)
[![Codecov test coverage](https://codecov.io/gh/r-heller/cellreportR/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-heller/cellreportR?branch=main)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/cellreportR)](https://cran.r-project.org/package=cellreportR)
[![CRAN downloads total](https://cranlogs.r-pkg.org/badges/grand-total/cellreportR)](https://cran.r-project.org/package=cellreportR)
[![License: MIT](https://img.shields.io/badge/license-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4
)
library(cellreportR)
```

cellreportR takes you from segmented single-cell microscopy data to a
structured, publication-ready diagnostic report. This vignette walks
through a full five-minute workflow using bundled synthetic data.

## 1. Load example data

The function `cr_example_experiment()` generates a realistic 96-well
synthetic experiment with six treatment groups, four channels, and
plate-edge / debris / contamination artefacts to exercise the QC
pipeline.

```{r example}
exp <- cr_example_experiment(seed = 42, n_cells_per_well = 60)
exp
```

## 2. Inspect the design

```{r design}
head(exp$design)
summary(exp)
```

## 3. Apply quality control

```{r qc}
exp_qc <- exp |>
  cr_qc_filter(min_area = 50, max_area = 5000, min_circularity = 0.2) |>
  cr_qc_doublets(k = 2.5)
cr_qc_summary(exp_qc)
```

## 4. Compute fold changes and run tests

```{r analysis}
res <- cr_test_all(exp_qc,
                   channel = "marker_1",
                   control_group = "Untreated",
                   level = "replicate")
attr(res, "summary")
```

## 5. Visualize

```{r plots}
cr_plot_intensity(exp_qc, "marker_1")
cr_plot_effect_sizes(res)
cr_plot_foldchange(res)
```

## 6. Logistic regression and ROC

```{r logistic}
logit <- cr_logistic(exp_qc,
                     channel = "marker_1",
                     treatment = "CompoundA_high",
                     control = "Untreated")
cr_plot_roc(logit)
```

## 7. Next steps

- See `vignette("statistical-analysis")` for details on the choice
  between cell-level and replicate-level tests.
- See `vignette("dose-response")` for IC50/EC50 fitting.
- Run `cr_run_app()` to explore your data interactively.

# Use of LLM tools

Portions of this package were prepared with assistance from large language model tooling for
narrowly defined, non-authorial tasks: copyediting, prose smoothing, Markdown/LaTeX formatting,
scaffolding of boilerplate files (CI configs, build scripts), code refactoring. The tools used were [Chat AI](https://kisski.gwdg.de/leistungen/2-02-llm-service/),
the LLM service of KISSKI (GWDG), and a self-hosted **Mistral Small (24B, Apache-2.0)** run locally via
[Ollama](https://ollama.com/) and the `ollamar` R package — local inference only, with no data sent to
third parties for the self-hosted model.

All scientific claims, methodological choices, analyses, interpretations, and conclusions are the
author's own. No LLM-generated text was incorporated without review and revision, and every reference
was verified against its DOI, arXiv ID, or ISBN.
