---
title: "Configurable Laboratory Reports with cellreportR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configurable Laboratory Reports with cellreportR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>")
```

The laboratory-reporting layer separates analytical data and QC calculations
from report metadata, classification, interpretation, rendering, and technical
provenance. Rendering consumes finalized objects and does not recalculate or
reinterpret analytical values.

```{r example}
exp <- cellreportR::cr_example_experiment(seed=42,n_cells_per_well=3)
spec <- cellreportR::cr_report_spec(
  report=list(title="Laboratory Report",report_id="EXAMPLE-001",version="1.0",status="DRAFT"),
  laboratory=list(name="Example Laboratory"),
  subject=list(subject_id="SUBJECT-001"),
  specimen=list(specimen_id="SPECIMEN-001",specimen_type="Example specimen"),
  examination=list(name="Example fluorescence assay",assay_version="1.0"),
  result=list(value=1.42,unit="a.u.",classification="HIGH",qc_status="PASS",source="manual"),
  interpretation=list(summary="The result falls within the configured upper category.",text="Interpretation text supplied by the user."),
  limitations="Example limitation supplied by the user.",
  authorization=list(reviewed_by="Example Reviewer",authorized_by="Example Authorizer")
)
qc <- cellreportR::cr_report_qc("Run acceptance","met","Configured criteria","PASS")
report <- cellreportR::cr_lab_report(exp,spec,qc)
cellreportR::cr_report_data(report)$spec$result
```

`cr_validate_report_spec()` reports `ERROR`, `WARNING`, and `INFO`-compatible
issues. Laboratories can add dotted paths through `required_fields` without
baking institutional policy into the package. `AMENDED` reports record their
superseded report and reason; the package represents this metadata but is not a
document-management or authentication system.

Use `cr_export_report_spec()` for schema-versioned interchange and
`cr_export_report_audit()` for concise provenance. The audit record includes
package, R, schema, and template versions, structured-data hashes, selected QC,
and experiment summaries without copying the full cell table. Report fields
may contain confidential information; they are not printed in routine console
messages, placed in filenames, or copied to the Shiny analysis log.

```{r shiny, eval=FALSE}
cellreportR::cr_run_app(experiment=exp,report_spec=spec)
```

The **Laboratory report** tab builds the same `cr_report_spec`, validates it,
previews the result, and downloads PDF, audit JSON, or specification JSON.
Scientific thresholds and interpretation remain user inputs. These features
are designed to support traceability and can form part of a laboratory-
controlled workflow; they do not guarantee regulatory or accreditation
compliance.
