---
title: "Configurable Laboratory Reports with cellreportR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configurable Laboratory Reports with cellreportR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>")
```

The laboratory-reporting layer separates analytical data and QC calculations
from report metadata, classification, interpretation, rendering, and technical
provenance. Rendering consumes finalized objects and does not recalculate or
reinterpret analytical values.

```{r example}
exp <- cellreportR::cr_example_experiment(seed=42,n_cells_per_well=3)
spec <- cellreportR::cr_report_spec(
  report=list(title="Laboratory Report",report_id="EXAMPLE-001",version="1.0",status="DRAFT"),
  laboratory=list(name="Example Laboratory"),
  subject=list(subject_id="SUBJECT-001"),
  specimen=list(specimen_id="SPECIMEN-001",specimen_type="Example specimen"),
  examination=list(name="Example fluorescence assay",assay_version="1.0"),
  result=list(value=1.42,unit="a.u.",classification="HIGH",qc_status="PASS",source="manual"),
  interpretation=list(summary="The result falls within the configured upper category.",text="Interpretation text supplied by the user."),
  limitations="Example limitation supplied by the user.",
  authorization=list(reviewed_by="Example Reviewer",authorized_by="Example Authorizer")
)
qc <- cellreportR::cr_report_qc("Run acceptance","met","Configured criteria","PASS")
style <- cellreportR::cr_report_style(
  paper="A4",mode="colour",density="standard",locale="en"
)
report <- cellreportR::cr_lab_report(exp,spec,qc,style=style)
cellreportR::cr_report_data(report)$spec$result
```

The human-facing PDF uses laboratory-report template 2.0. The package-owned
KOMA-Script layout places report identity, subject/specimen summary, the
examination, the primary result, and its interpretation on the first page for
normal content. Quality control, traceability, limitations, additional
information, and authorization form the second-page release record. Every page
shows report ID, version, status, and `Page X of Y`; the machine-readable report
schema remains independently versioned at 1.0.

`cr_report_display_data()` exposes the exact human labels and formatted values
used for rendering while keeping stored ISO datetimes unchanged. A reusable
profile can supply laboratory branding and policy once:

```{r profile}
profile <- cellreportR::cr_report_profile(
  laboratory=list(name="Example Laboratory",department="Analytical Services"),
  style=cellreportR::cr_report_style(
    mode="grayscale",locale="de",show_signature_lines=FALSE
  ),
  labels=c(specimen_id="Sample ID")
)
profile$style
```

Colour and grayscale are intentional document modes rather than post-render
conversions. In either mode, classification and QC states remain explicit text.
PDF rendering requires Pandoc and XeLaTeX; `cr_render_lab_report()` checks the
engine, standard TeX components, logo, template, and output location before
rendering. The technical appendix is excluded by default because audit JSON is
the preferred machine-readable companion; opt in with
`cr_report_style(include_audit_appendix=TRUE)` when a visible appendix is
needed.

`cr_validate_report_spec()` reports `ERROR`, `WARNING`, and `INFO`-compatible
issues. Laboratories can add dotted paths through `required_fields` without
baking institutional policy into the package. `AMENDED` reports record their
superseded report and reason; the package represents this metadata but is not a
document-management or authentication system.

Use `cr_export_report_spec()` for schema-versioned interchange and
`cr_export_report_audit()` for concise provenance. The audit record includes
package, R, schema, and template versions, structured-data hashes, selected QC,
and experiment summaries without copying the full cell table. Report fields
may contain confidential information; they are not printed in routine console
messages, placed in filenames, or copied to the Shiny analysis log.

```{r shiny, eval=FALSE}
cellreportR::cr_run_app(experiment=exp,report_spec=spec)
```

The **Laboratory report** tab builds the same `cr_report_spec`, validates it,
previews the result, and downloads PDF, audit JSON, or specification JSON.
Scientific thresholds and interpretation remain user inputs. These features
are designed to support traceability and can form part of a laboratory-
controlled workflow; they do not guarantee regulatory or accreditation
compliance.
