---
title: "Statistical Analysis of Cell-Based Assays"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Analysis of Cell-Based Assays}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-heller/cellreportR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-heller/cellreportR/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/r-heller/cellreportR/actions/workflows/pkgdown.yaml/badge.svg)](https://r-heller.github.io/cellreportR/)
[![CRAN status](https://www.r-pkg.org/badges/version/cellreportR)](https://CRAN.R-project.org/package=cellreportR)
[![Codecov test coverage](https://codecov.io/gh/r-heller/cellreportR/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-heller/cellreportR?branch=main)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/cellreportR)](https://cran.r-project.org/package=cellreportR)
[![CRAN downloads total](https://cranlogs.r-pkg.org/badges/grand-total/cellreportR)](https://cran.r-project.org/package=cellreportR)
[![License: MIT](https://img.shields.io/badge/license-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cellreportR)
```

## Why hierarchical testing matters

In a microscopy-based cell assay, each treatment condition is measured
across several wells, and each well contains many individual cells.
Cells are not independent: their fluorescence is correlated within a
well (shared staining, focus, exposure). Treating every cell as an
independent replicate inflates sample size and gives optimistic
p-values.

cellreportR supports two standard resolutions:

- **Cell-level** tests treat each cell as a data point. Use these for
  distributional comparisons (e.g. is there a subpopulation shift).
- **Replicate-level** tests aggregate cells to one number per well,
  then test wells. Use these for the primary inference about whether
  a compound has an effect.

A defensible workflow is to run both and report effect size with CI
alongside p-values at the replicate level.

## A worked example

```{r workflow}
exp <- cr_example_experiment(seed = 2, n_cells_per_well = 60)
res <- cr_test(exp, "marker_1", "CompoundA_high", "Untreated",
               test = "mann_whitney", level = "both")
res$cell_level
res$rep_level
res$effect_sizes
```

Notice that the cell-level p-value is much smaller than the
replicate-level p-value — a consequence of the inflated sample
size.

## Effect size interpretation

cellreportR computes Cohen's d, Hedges' g, Cliff's delta and
rank-biserial correlation. Conventional benchmarks:

- |d| < 0.2 — negligible
- 0.2 ≤ |d| < 0.5 — small
- 0.5 ≤ |d| < 0.8 — medium
- |d| ≥ 0.8 — large

Cliff's delta benchmarks are 0.147 / 0.33 / 0.474.

```{r effectsize}
set.seed(1)
cr_effect_size(rnorm(100, 1), rnorm(100, 0))
```

## ROC / AUC for discriminability

When the question is "does this marker discriminate treated from
control cells?", fit a univariate logistic regression:

```{r roc}
logit <- cr_logistic(exp, "marker_1", "CompoundA_high", "Untreated")
cr_auc(logit)
cr_plot_roc(logit)
```

## Multiple testing correction

`cr_test_all()` applies `stats::p.adjust()` by default with the
Benjamini-Hochberg method:

```{r adjust}
all_res <- cr_test_all(exp, "marker_1", "Untreated",
                       level = "replicate")
attr(all_res, "summary")
```

## Power considerations

A quick hierarchical-aware power check:

```{r power}
cr_power_analysis(effect_size = 0.8,
                  n_replicates = 4,
                  n_cells_per_rep = 100,
                  n_sim = 200)
```
